Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-05-03 doc/paper.Rmd
fix citations

Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [***(CITATION Bruna, Gaoue)***] We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
This time series, which included the severe droughts of 1997 [@mcphadenChildProdigy1997981999] and 2005 [@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008], allowed us to address the following questions:\

@@

\< !--
\# Gaoue = ?
Orou G. Gaoue, Carol C. Horvitz, Ulrich K. Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,<https://doi.org/10.1016/j.ecolmodel.2019.03.007>.
-- \>

@@

    -   low precip = trees lose leaves or die, more light gets through or gaps open.
        This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses20

@@

<!--# in intro, questions are 1: what are effects on vital rates, 2: are effects delayed, 3: differences between habitats.  Keep the same order here? -->

@@

Instead, we believe that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on others species.
For example, the mortality rate of tropical trees can spike in the year following a severe drought.
The growth and flowering rates of understory plants would only increase years later, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls increase light penetration to the forest understory (**CITATIONS)**.
Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments <!--#  (question about wording - does this mean the delay is relatively greater, or the magnitude of the effect is greater?)-->.
Unfortunately, experiments disentangling the complex ways in which light, environmental conditions, and ecological neighborhood influence temporal variation in vital rates are rare (sensu @westerband2015, @westerbandEarlyLifeConditions2017, @horvitzEffectsPlantSize2002, @gagnon).
Without such experiments, generalizations about whether delayed effects are driven by direct or indirect mechanisms will continue to prove elusive.

>>>>

Herbaceous plants represent up to 25% of plant diversity in tropical forests [@gentryContributionNontreesSpecies1987], are critical food and habitat for myriad species [@snowTropicalFrugivorousBirds1981], and are economically and culturally vital [@nakazonoExperiementalHarvestingNontimber2004; @ticktinRelationshipsNinoSouthern2003].
Nevertheless, the impacts of global change phenomena on their demography remain conspicuously understudied [@brunaHabitatFragmentationPlant2009] <!--# Emilio had written "Gaoue" as potential reference, but the on I have in Zotero is about trees, not understory plants -->.
We used a decade of demographic and climatic data from an experimentally fragmented landscape in the Central Amazon to assess the effects of climate on populations of a tropical understory herb (*Heliconia acuminata*, Heliconiaceae).
This time series, which included the severe droughts of 1997 [@mcphadenChildProdigy1997981999] and 2005 [@marengoDroughtAmazonia20052008; @zengCausesImpacts20052008], allowed us to address the following questions:\

@@

Gaoue = ?
Orou G. Gaoue, Carol C. Horvitz, Ulrich K. Steiner, Shripad Tuljapurkar, Climate, rather than human disturbance, is the main driver of age-specific mortality trajectories in a tropical tree, Ecological Modelling, Volume 400, 2019, Pages 34-40,<https://doi.org/10.1016/j.ecolmodel.2019.03.007>.

-   Eric: I don't think so. It was a reference supporting that understory plants are understudied.

@@

    -   low precip = trees lose leaves or die, more light gets through or gaps open.
        This can be positive (more light = more growth) but not always ( @westerbandInteractionsPlantSize2015) there is a threshold that can lead to photoinhibition ( @hePhotoinhibitionHeliconiaNatural1996, @mulkeyComparativeLifeHistory1991).
        Drought and herbivory effects can be additive, at least in seedlings ( @bartonAdditiveNonadditiveResponses2020 )

@@

@@

Instead, we believe that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on others species.
For example, the mortality rate of tropical trees can spike in the year following a severe drought.
The growth and flowering rates of understory plants would only increase years later, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls increase light penetration to the forest understory (**CITATIONS)**.
Drought-driven litterfall is also greater at forest edges than forest interiors (**CITATIONS)**, which could explain the increased delayed effects of SPEI on survival and growth in forest fragments <!--#  (question about wording - does this mean the delay is relatively greater, or the magnitude of the effect is greater?)-->.
Unfortunately, experiments disentangling the complex ways in which light, environmental conditions, and ecological neighborhood influence temporal variation in vital rates are rare [sensu @westerband2015; @westerbandEarlyLifeConditions2017; @horvitzEffectsPlantSize2002; @gagnonGrowthUnderstoryHerb2011]<!--# what is this "sensu" in reference to?  The definition of vital rates?  of "rare"? --> .
Without such experiments, generalizations about whether delayed effects are driven by direct or indirect mechanisms will continue to prove elusive.
